---
title: "R Notebook"
output: html_notebook
---
## Loading libraries and load clean data
```{r load_library}
suppressPackageStartupMessages(library(DESeq2))
suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(reshape))
```
Load and process `sanitized.dds` - renamig and adding column data.
```{r load_data}
pkgDir <- "/fh/fast/tapscott_s/CompBio/RNA-Seq/hg38.FSHD.biopsyI-IV"
source(file.path(pkgDir, "scripts", "viz_tools.R"))
source(file.path(pkgDir, "scripts", "renameSample.R"))
dataDir <- file.path(pkgDir, "data")
figDir <- file.path(pkgDir, "manuscript", "figures")
load(file.path(pkgDir, "data", "sanitized.dds.rda"))
load(file.path(pkgDir, "data", "mk.rda"))
load(file.path(pkgDir, "data", "fourmk.rda"))
source(file.path(pkgDir, "scripts", "renameSample.R"))
new_sampleID <- renameSample(colnames(sanitized.dds))
sanitized.dds$new_sampleID <- new_sampleID
colnames(sanitized.dds) <- new_sampleID
sanitized.dds$RNA.Group <- 
  paste0("G", sanitized.dds$RNA.Cluster,
         "_", sanitized.dds$pheno_type)
sanitized.dds$RNA.Group <- 
  factor(sanitized.dds$RNA.Group)
rlg <- rlog(sanitized.dds, 
            blind=TRUE)
```

## Defind dold group
cold group = G1_FSHD and Control
```{r define_cold_group}
#' sample subset - Group1 including control and some FSHD
cold_group <- sanitized.dds$RNA.Cluster %in% c(1)
```

## Questions
- Can we separate the G1_FSHD and G2_FSHD samples from the controls?

- If yes, how many dimenstion of PCA is appropriate? and which genes are responsible for the separation??pr


## PCA

### Feature selection
Pre-select top one-thired genes with most vairation.

```{r pre_selction}
pre_selection <- function(
  method=c("by_DESeq", "by_rowVars"), ntop=500, object) {
  if (method == "by_DESeq") {
    dds <- DESeq(object)
    res <- results(dds)
    res <- res[order(res$padj), ]
    pre_selected <- rownames(res)[1:ntop]
  }
  
  if (method == "by_rowVars") {
    rv <- rowVars(assay(object))
    idx <- 
      order(rv, decreasing=TRUE)[seq_len(min(ntop,
                                             length(rv)))]
    pre_selected <- rownames(object)[idx]
    
  }
  pre_selected
}
```

```{r do_selection}
pre_sel <- pre_selection(ntop=1000, method="by_rowVars",
                         object=rlg[, cold_group])
```

```{r simple_pca}
sub_rlg <- rlg[pre_sel, cold_group]
pca <- prcomp(t(assay(sub_rlg)))
percentVar <- pca$sdev^2 / sum( pca$sdev^2 )

#' arrange the returned pca data
library(RColorBrewer)
myPal <- brewer.pal(4, "Set2")
myPal <- c("pink", "green")
myCol <- myPal[factor(sub_rlg$RNA.Group)]
pairs(pca$x[, 1:4], col=myCol)
```

```{r show_pca_var}
#' umsum(percentVar)
plot(cumsum(percentVar), type="l",
     xlab="number of PC",
     ylab="cummulative variance")
```

### Importand features
Identify the top 100 impartant features that maximize the variance among samples.

```{r important_features}
library(pheatmap)
loadings <- abs(pca$rotation)
getNTop <- function(pc_loading, name="PC1", ntop=10) {
  ord <- order(pc_loading[, name], decreasing=TRUE)
  rownames(pc_loading[ord[1:ntop], ])
}

ntop <- 100
ntop_df <- data.frame(PC1=getNTop(loadings, "PC1", ntop),
                      PC2=getNTop(loadings, "PC2", ntop))

ntop_features <- unique(union(ntop_df$PC1, ntop_df$PC2))
```

Prepare data for plotting heatmap.
```{r viz_heatmap}
data <- assay(sub_rlg[ntop_features, ])
rownames(data) <-
  rowData(sub_rlg[ntop_features])$gene_name
annotation_col <- data.frame(RNA.Group=sub_rlg$RNA.Group)
rownames(annotation_col) <- colnames(sub_rlg)
pheatmap(data, 
         annotation_col=annotation_col,
         fontsize_row=4, 
         show_rownames=FALSE,
         clustering_distance_rows = "manhattan",
         clustering_distance_cols = "manhattan",
         scale="row")
```

Output the heatmap to a PDF file.
```{r output_import_feature}
pheatmap(data, 
         annotation_col=annotation_col,
         fontsize_row=2, 
         show_rownames=TRUE,
         scale="row",
         silent=TRUE,
         clustering_distance_rows = "manhattan",
         clustering_distance_cols = "manhattan",
         filename=file.path(pkgDir, "figures",
           "heatmap_pca_loading_top100.pdf"))
```
## Sparse PCA
Use the sparse PCA to find the most important features that differentiate Control and the G1_FSHD outliers.
```{r session_info}
sessionInfo()
```
